Random Walks across Social Networks: The Markovian Movement of a Stochastic Soccerball

### 2020-07-20 rmd/draft.Rmd
just need to write results and conclusions, but i've used up wordcount

@@

@RPollardReep1997
@RPollard2002
@RPollardEtAl1988
@CReepBenjamin1968
@CReepEtAl1971

even earlier @JEaves2015

David Barrett [@ECochraneEtAl1910]

@@

dataset
@LPappalardoEtAl2019a

nested JSON

- load JSON
- decode the 'tags', player IDs and team IDs.
- unpack the location and target positions.
- distinguish play sequences, as indicated by a 'Free Kick' (includes Throw-in) or a 'Goal'.
- distinguish possession sequences
- include starting XI and Home/Away status info
- convert time-stamp in seconds to classic minutes

load into R, test that league table given by preprocessed data matches actual results. Table 1
- found that De Bruyne's goal was misclassified as an inaccurate pass, rather than a goal-scoring shot.

Table 1
Premiership Top 5

Initial exploratory visualizations:

- points showing events in a match for home and away team

Figure 1a,b
<img src="../figures/EventMapHome.png">

Figure 2a,b
- arrows showing passes and shots
<img src="../figures/PassMapHome.png">

More specifically: we consider a specific possession sequence to check that our preprocessing strategy has been successfuly.
Table 2: Events of Possession Sequence
Figure 3: Pitchmap showing Possession Sequence
<img src="../figures/PossessionSequence.png">

Figure 4: YouTube Clip showing successful conclusion of possession sequence
<img src="../figures/youtube_clip.png">

@@

*Social network analysis* uses the tools of mathematical graph theory to analyse social structures. We therefore begin by defining several necessary theoretical terms.

@@

@@

There are then a large variety of different graph metrics which can be calculated, including in particular an ever-increasing number of measures of *centrality*.

@@

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass. This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy. However, visualizing the mean position of a player could be misleading: if a player always plays on the wing, but is switched half-way through the game from one side of the pitch to the other, his 'average position' will be central, although he may never have touched the ball centrally. It is also not possible in a static image to distinguish a pass $e_{U,V}$ from its reverse $e_{V,U}$. @KSingh2018 therefore suggests creating interactive visualizations in which the user can adjust the weighting to highlight forward or backwards passes as desired.

Figure 7
<img src="../figures/PassingNetworkHome.png">

@@

We then also need some definitions from *probability theory* and *statistics*. [@RSerfozo2009a]

@@

A Markov process is *time-homogenous* if the transition probabilities stay constant through time, ie. $\mathbb{P}(X_{n+1}=s|X_{n}) = \mathbb{P}(X_{m+1}=x|X_{m}) \forall m,n \in T$. If $S$ is countable, then the probability distribution of such a process can be described by its initial *distribution vector* $\pi_{0}$ and its *transition matrix* $P$, where each $i$th row is made up of the transition probabilities $p_{ij}$ and $\sum\limits_{j} p_{ij} = 1 \forall i$. Specifically, after $k$ transitions we have the distribution $\pi_{k} = \pi_{0}P^{k}$.

If a Markov process develops in discrete time and takes a values from a discrete state space, then we call it a *Markov chain* and can straightforwardly draw it as a weighted simple graph by treating its transition matrix as an adjacency matrix. Conversely, we can transform an adjacency matrix $M$ into a transition matrix $P$ by defining the elements $p_{ij} := \frac{m_{ij} }{\sum\limits_{j}m_{ij}}$, so that each row sums to 1 as required. We can then picture the development of a Markov process as a *random walk* on the graph defined by its transition probabilities, noting that since the transition probabilities are not uniform this walk is *biased*.

@@

Given an initial distribution $\pi_{0}$, we know that after $k$ transitions, the state $\pi_{k} = \pi_{0}P^{k}$.

We can define the *state vector* $\delta$ of any particular state $s{i}$ as
$$
\delta_{j} := \begin{cases}
    \begin{array}{ll}
    	1, & i=j \\
	0, & \mathrm{otherwise}
    \end{array}
\end{cases}
$$

The probability $\mathbb{P}(X_{k}=s)$ is then given by the dot product $\langle \pi_{k}, \delta \rangle$.

If $s_{i}$ is an absorbing state, then the probability that it is reached *for the first time* is given by  $p_{k}:= \langle \pi_{k}, \delta \rangle - \langle \pi_{k-1}, \delta \rangle$.

If the expectation of the hitting time is finite it can then be approximated directly by iterating up to some large $K$:
$$\mathbb{E}[\tau_{s}] = \sum_{k=0}^\infty kp_k \approx \sum_{k=0}^K kp_k$$

Thus we can quantify, for each player on the team, the expected number of events there will be before the team scores a goal. We invert this to give a measure of the expected number of goals; and scale it by the team's average number of events per game, and by the average player's expected goals values, and call this measure *Markov Expected Goals* (MXG).

@@

>>>>

Association football is the world's most popular game and a multibillion dollar industry. Statistical analysis of the sport was pioneered more than fifty years ago, but has only recently begun to make a more widespread impact. In this paper I explain how Social Network Analysis can provide the tools to model a football team's ball-possession as a dynamic probabilistic system, and hence how we can quantify the effect of each player on the team's goal-scoring capability.

@@

After thirty years of waiting, Liverpool F.C. are once again English champions [@BBC2020]. But while manager Jurgen Klopp is now guaranteed his place in footballing folklore, less well known is the name of data scientist Ian Graham, whose analysis led to Klopp being hired in the first place [@BSchoenfeld2019], and whose research team is regarded as "the strongest of its kind in English football" [@SAustin2020].

In this paper, I develop a rigorous account of a football match as a stochastic social network, the graph of which is generated by the movement of the ball from player to player. I then show how the adjacency matrix of a player, considered as a node in a football match's possession sequence graph, can be used to quantify his contribution to the likelihood of the team scoring. Finally, I discuss the potential and pitfalls of this new metric, which I call *Relative Markovian Expected Goals*.

@@

### JSON Event-logs

The dataset is provided by WyScout, one of a growing number of companies in the football analytics industry, and is available from FigShare.com [@LPappalardoMassucco2020]. It includes more than three million *event-logs* for every match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

### Preprocessing

I used Python to convert this into a *tidy* dataframe [@HWickham2014], that could then be analyzed and visualized using R. For the sake of reproducible research, all code is attached as appendices, as well as available on Github.

As part of this preprocessing stage I also labelled each event according first to its *in-play sequence*, and then its *possession sequence*. An in-play sequence ends when the ball goes out of play, as indicated by the subsequent event having the *eventName* 'Free Kick' (a category which includes throw-ins and corners, as well as actual free-kicks given for fouls or offsides); or by the final two events (the shot and the attempted save) being labelled with the *tag* 'Goal'. (Although every goal is followed by a 'free kick' from the centre of the pitch, this never seems to be included in the event logs, presumably because they are created from camera footage which focusses on the goal celebrations and tends to miss the immediate restart of play after a goal).

Intuitively, to identify a possession sequence we should just be able to iterate through the event-logs of a given in-play sequence, and note when the team of the player changes. However, since the event-logs include challenges for the ball even when they are unsuccessful, this would lead to systematically undercounting the length of possession sequences. The solution I adopt is to iterate through the event-logs for a single team within an in-play sequence, paying close attention to the location data: if the location of an event is identical to the target-location of the previous event associated with that same team, then I treat is as part of the same possession sequence, otherwise I treat it as a separate sequence.

### Exploratory Visualization

We can then visualize the location of the event-log data points in the context of a diagrammatic football pitch, to allow for an immediate and accessible view of any particular possession sequence (eg. Table 2, Figure 3, Figure 4).

Figure 3: Pitchmap showing Possession Sequence
<img src="../figures/PossessionSequence.png">

We can compare this to a tabulated account of all events from just before the beginning of the possession sequence until its end, we see that our algorithm has successfully handled the case where an opposition player's unsuccessful challenge interrupts the sequence of events associated with the team in possession.

Table 2: Events of Possession Sequence

And with possession sequences that end in a goal, it is generally possible to find video highlights of the events under consideration, and confirm that the data we have does correspond to what actually happened.

Figure 4: YouTube Clip showing successful conclusion of possession sequence
<img src="../figures/youtube_clip.gif">

@@

Social network analysis applies the concepts of mathematical *graph theory*.

@@

On an undirected graph, we call the number of edges connecting a node its *degree*; in a directed graph we distinguish between the *indegree* and *outdegree*.

@@

@@

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass. This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.
Figure 7
<img src="../figures/PassingNetworkHome.png">

@@

To consider the team's ball-possession in its dynamic development, rather than merely as a static structure, we also need some definitions from *probability theory*.

@@

A Markov process is *time-homogenous* if the transition probabilities stay constant through time, ie. $\mathbb{P}(X_{n+1}=s|X_{n}) = \mathbb{P}(X_{m+1}=x|X_{m}) \forall m,n \in T$. If $S$ is countable, then the probability distribution of such a process can be described by its initial *distribution vector* $\boldsymbol{\alpha_{0}}$ and its *transition matrix* $\mathbf{P}$, where each $i$th row is made up of the transition probabilities $p_{ij}$ and $\sum\limits_{j} p_{ij} = 1 \forall i$. Specifically, after $k$ transitions we have the distribution $\boldsymbol{\alpha_{k}} = \boldsymbol{\alpha_{0}}\mathbf{P}^{k}$.

If a Markov process develops in discrete time and takes a values from a discrete state space, then we call it a *Markov chain* and can straightforwardly draw it as a weighted simple graph by treating its transition matrix as an adjacency matrix. Conversely, we can transform an adjacency matrix $\mathbf{M}$ into a transition matrix $\mathbf{P}$ by defining the elements so that each row sums to 1 as required:

$p_{ij} := \frac{m_{ij} }{\sum\limits_{j}m_{ij}}$

We can then picture the development of a Markov process as a *random walk* on the graph in which the nodes ${V}$ of the graph are the state space $S$ of the process, and the weights of the edges are the transition probabilities.

@@

Given an initial distribution $\boldsymbol{\alpha_{0}}$, we know that after $k$ transitions, the state $\boldsymbol{\alpha_{k}} = \boldsymbol{\alpha_{0}}\mathbf{P}^{k}$.

We can define the *state vector* $\boldsymbol{\delta_{i}}$ of any particular state $s_{i}$ as
$$
\delta_{ij} := \begin{cases}
    \begin{array}{ll}
    	1, & i=j \\
	0, & \mathrm{otherwise.}
    \end{array}
\end{cases}
$$

The probability $\mathbb{P}(X_{k}=s_{i})$ is then given by the dot product $\langle \boldsymbol{\alpha_{k}}, \boldsymbol{\delta_{i}} \rangle$.

If $s_{i}$ is an absorbing state, then the probability that it is reached *for the first time* is given by $\pi_{k}:= \langle \boldsymbol{\alpha_{k}}, \boldsymbol{\delta_{i}} \rangle - \langle \boldsymbol{\alpha_{k-1}}, \boldsymbol{\delta_{i}} \rangle$.

If the expectation of the hitting time is finite it can then be approximated directly by iterating up to some large $K$:
$$\mathbb{E}[\tau_{s}] = \sum_{k=0}^\infty k\pi_k \approx \sum_{k=0}^K k\pi_k$$

Thus we can quantify, for each player on the team, the expected number of events there will be before the team scores a goal. We invert this to give a measure of the expected number of goals; and scale it by the team's average number of events per game, we call this measure *Markovian Expected Goals* (MXG). We then scale this by the team's mean expected goals values to give a *Relative MXG* value, so that values can be compared fairly across teams to see which individual players contribute the most to their teams' goalscoring chances.

@@

T
